---
title: "HW 2: Diabetes"
description: |
 This document looks into certain factors that may contribute to someone developing diabetes.
author:
  - name: Peter Sullivan 
   
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include= TRUE}
library(tidyverse)

diabetes_data <- read.csv("diabetes.csv", sep = ',')
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
head(diabetes_data)
colnames(diabetes_data)
dim(diabetes_data)
```


#
When looking at the Diabetes data set, there are `r nrow(diabetes_data)` observations, and `r ncol(diabetes_data)` variables. This data set was posted on kagle.com as a machine learning exercise, to see if someone can predict whether the patient has diabetes based on certain features. The target column, "Outcome", states whether someone has diabetes; 1 for diabetes and 0 for not having diabetes.

Here are the column names: `r colnames(diabetes_data)`

```{r , include=TRUE}
diabetes_data %>%
ggplot() +
  geom_point(mapping = aes(x = Age, y = BloodPressure))+
  geom_smooth(mapping = aes(x = Age, y = BloodPressure))


```


#
Here is an attempt at using a Facet wrap, specifically looking at the number of pregnancies per patient:


```{r, include = TRUE}
diabetes_data %>%
ggplot() +
  geom_point(mapping = aes(x = Age, y = BloodPressure)) +
  geom_smooth(mapping = aes(x = Age, BloodPressure)) +
  facet_wrap( ~ Pregnancies , nrow = 2)
```

#
This is another example of using a Facet wrap but instead of looking at pregnancies, we are looking at the Diabetes Outcome.

```{r, include = TRUE}

ggplot(data = diabetes_data) +
  geom_point(mapping = aes(x = Age, y = BloodPressure)) +
  geom_smooth(mapping = aes(x = Age, BloodPressure)) +
  facet_wrap( ~ Outcome , nrow = 2)
```


Distill is a publication format for scientific and technical writing, native to the web. 

Learn more about using Distill for R Markdown at <https://rstudio.github.io/distill>.





